---
title: "Basic analysis of the affects of n_points on overlaps and delta metrics"
author: "Joe DiRocco"
date: "April 10, 2026"
output: html_document
---

# Observations

It is made clear through the visualizations that as n_points increases so does
the probability of an overlapping. Extreme outliers were observed more often as n_points increased leading to heavy skewness in the delta metrics. 
Our previous understanding of the overlap risk of these unfolding methods has held true. At n_points = 100 DFS has a 7% overlap rate, shortest_paths nearly 1%, and bfs with 0.4%. 


```{r include=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(tidyr)
  library(readr)
  library(broom)
})

data_path <- "unfolding_method_metrics.csv"

df <- read_csv(data_path, show_col_types = FALSE) %>%
  mutate(
    method = as.factor(method),
    n_points = as.integer(n_points),
    has_overlap = num_overlappings > 0,
    overlap_rate = num_overlappings / pairs,
    neg_min_delta = min_delta < 0
  )

cat("Rows:", nrow(df), "\n")
cat("Methods:", paste(sort(unique(df$method)), collapse = ", "), "\n")
cat("n_points:", paste(sort(unique(df$n_points)), collapse = ", "), "\n\n")
```

# Summary by n_points only

```{r, echo=FALSE, warning=FALSE}
summary_by_n <- df %>%
  group_by(n_points) %>%
  summarise(
    trials = n(),
    overlap_prob = mean(has_overlap),
    mean_num_overlappings = mean(num_overlappings),
    mean_min_delta = mean(min_delta),
    median_min_delta = median(min_delta),
    mean_mean_delta = mean(mean_delta),
    median_mean_delta = median(mean_delta),
    mean_median_delta = mean(median_delta),
    .groups = "drop"
  )

knitr::kable(summary_by_n)
```


# Summary by method and n_points
```{r, echo=FALSE, warning=FALSE}
summary_by_method_n <- df %>%
  group_by(method, n_points) %>%
  summarise(
    trials = n(),
    overlap_prob = mean(has_overlap),
    mean_num_overlappings = mean(num_overlappings),
    mean_min_delta = mean(min_delta),
    median_min_delta = median(min_delta),
    mean_mean_delta = mean(mean_delta),
    median_mean_delta = median(mean_delta),   # add this
    mean_median_delta = mean(median_delta),
    median_median_delta = median(median_delta), # add this
    neg_min_delta_prob = mean(neg_min_delta),
    .groups = "drop"
  )

knitr::kable(summary_by_method_n)
```

# Tail-risk comparison for min_delta

This compares the extreme lower tail (worst 5%) to a less-extreme tail band
(15-20th percentile, labeled here as the 80-85% worst band) to show how sharp
the drop becomes near the worst outcomes.

```{r, echo=FALSE, warning=FALSE}
tail_profile <- df %>%
  group_by(method, n_points) %>%
  summarise(
    q05 = quantile(min_delta, 0.05, na.rm = TRUE),
    q15 = quantile(min_delta, 0.15, na.rm = TRUE),
    q20 = quantile(min_delta, 0.20, na.rm = TRUE),
    band_80_85_mid = (q15 + q20) / 2,
    tail_drop = band_80_85_mid - q05,
    tail_drop_ratio = if_else(abs(band_80_85_mid) > 0, q05 / band_80_85_mid, NA_real_),
    .groups = "drop"
  )

knitr::kable(tail_profile)
```

```{r, echo=FALSE, warning=FALSE}
tail_profile_long <- tail_profile %>%
  select(method, n_points, q05, band_80_85_mid) %>%
  pivot_longer(
    cols = c(q05, band_80_85_mid),
    names_to = "tail_metric",
    values_to = "value"
  )

ggplot(tail_profile_long, aes(x = n_points, y = value, color = method, linetype = tail_metric)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "min_delta tail comparison: worst 5% vs 80-85% worst band",
    x = "n_points",
    y = "min_delta",
    color = "method",
    linetype = "tail metric"
  ) +
  theme_minimal()
```

Interpretation: `q05` shows the worst-case behavior in the lower tail, while
`band_80_85_mid` is a less extreme bad-case baseline. DFS has a sharp tail cliff and a small fraction of
trials become much worse than the rest.


```{r, include=FALSE}
# Overlap probability model
overlap_model <- glm(
  has_overlap ~ n_points + method,
  data = df,
  family = binomial()
)

tidy(overlap_model)

# Delta trend models
lm_min_delta <- lm(min_delta ~ n_points * method, data = df)
lm_mean_delta <- lm(mean_delta ~ n_points * method, data = df)
lm_median_delta <- lm(median_delta ~ n_points * method, data = df)

tidy(lm_min_delta)

tidy(lm_mean_delta)

tidy(lm_median_delta)
```

# Visualizations
```{r}

# Overlap probability by n_points and method
p_overlap_prob <- summary_by_method_n %>%
  ggplot(aes(x = n_points, y = overlap_prob, color = method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(limits = c(0, 0.15)) +
  labs(
    title = "Overlap Probability vs n_points",
    x = "n_points",
    y = "P(num_overlappings > 0)",
    color = "method"
  ) +
  theme_minimal()

p_overlap_prob

# Mean number of overlaps by n_points and method
p_overlap_count <- summary_by_method_n %>%
  ggplot(aes(x = n_points, y = mean_num_overlappings, color = method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Mean Number of Overlappings vs n_points",
    x = "n_points",
    y = "Mean num_overlappings",
    color = "method"
  ) +
  theme_minimal()

print(p_overlap_count)

# Delta trends (summaries)
delta_summary_long <- summary_by_method_n %>%
  select(method, n_points, mean_min_delta, mean_mean_delta, mean_median_delta) %>%
  pivot_longer(
    cols = c(mean_min_delta, mean_mean_delta, mean_median_delta),
    names_to = "metric",
    values_to = "value"
  )

p_delta_summary <- delta_summary_long %>%
  ggplot(aes(x = n_points, y = value, color = method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~metric, scales = "free_y") +
  labs(
    title = "Delta Metrics vs n_points",
    x = "n_points",
    y = "Metric value",
    color = "method"
  ) +
  theme_minimal()

p_delta_summary

# Median-based delta trends
delta_median_long <- summary_by_method_n %>%
  select(method, n_points, median_min_delta, median_mean_delta, median_median_delta) %>%
  pivot_longer(
    cols = c(median_min_delta, median_mean_delta, median_median_delta),
    names_to = "metric",
    values_to = "value"
  )

p_delta_median <- delta_median_long %>%
  ggplot(aes(x = n_points, y = value, color = method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~metric, scales = "free_y") +
  labs(
    title = "Median Delta Metrics vs n_points",
    x = "n_points",
    y = "Median value",
    color = "method"
  ) +
  theme_minimal()

p_delta_median

# Raw distributions for delta metrics
df_long <- df %>%
  select(method, n_points, min_delta, mean_delta, median_delta) %>%
  pivot_longer(
    cols = c(min_delta, mean_delta, median_delta),
    names_to = "metric",
    values_to = "value"
  )

p_delta_box <- df_long %>%
  ggplot(aes(x = factor(n_points), y = value, fill = method)) +
  geom_boxplot(outlier.alpha = 0.2) +
  facet_wrap(~metric, scales = "free_y") +
  labs(
    title = "Distribution of Delta Metrics by n_points",
    x = "n_points",
    y = "Value",
    fill = "method"
  ) +
  theme_minimal()

p_delta_box
```
